Covariate analysis of MSSM reprocessed counts (with CQN normalisation and covariates adjusted)

### 2017-04-15 code/Rmd/MSSM_REPROCESSED.Rmd
Modified MSSM markdown for April 2017 freeze

knit2synapse::knitToFolderEntity(file = "./MSSM_REPROCESSED.Rmd",
                                 parentId = "syn7554308",
                                 entityName = 'MSSM Reprocessed RNASeq (Normalised)')
```

@@

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn7554308'
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of MSSM reprocessed RNASeq data';

@@

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data}
# Get reprocessed sample counts for temporal cortex
COUNT_ID <- 'syn7994853';
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F)

@@

# Get picard metrics from synapse
METADATA.PICARD_ID <- 'syn8380379'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)

METADATA.PICARD$sampleIdentifier = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
  gsub('.accepted_hits.sort.coord/picard.CombinedMetrics.csv','',.)
colnames(METADATA.PICARD) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD))
colnames(METADATA.PICARD) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD))

@@

# Fix metadata
METADATA = METADATA %>%
  dplyr::mutate(AOD = gsub('\\+','',AOD),
                BrainRegion = factor(BrodmannArea,
                                     levels = c('BM10', 'BM22', 'BM36', 'BM44'),
                                     labels = c('BM10' = 'FP', 'BM22' = 'STG', 
                                                'BM36' = 'PHG', 'BM44' = 'IFG')),
                Diagnosis = 'NONE', 
                RIN2 = RIN^2)
METADATA$Diagnosis[METADATA$CDR <= 1] = 'ND'
METADATA$Diagnosis[METADATA$CDR == 2 | METADATA$CDR == 3] = 'MD'
METADATA$Diagnosis[METADATA$CDR >= 4] = 'SD'
levels(METADATA$batch)[levels(METADATA$batch) == ''] = 'NoBatch'

@@

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering}
primaryVariable <- "BrainRegion.Diagnosis"
FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrainRegion.Diagnosis")
ContCovariates <- c("RIN", "RIN2", "AOD","PMI",  "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")

@@

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], 
                                                 MIN_GENE_CPM=0, 
                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

### Normalisation (with NULL)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

@@

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=5, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Diagnosis, size=RIN, shape = SEX))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(.~Region, scales = 'free_y')
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data1.2, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```

### Sanity check
```{r residual.adj, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

@@

# Plot first 4 PCs
plotdata <- data.frame(SampleID=rownames(PC$rotation), 
                       PC1=PC$rotation[,1], 
                       PC2=PC$rotation[,2])

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Diagnosis, size=RIN, shape = SEX))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(.~Region, scales = 'free_y')
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```

@@

Interpretation of BrainRegion.Diagnosis
1. FP.MD: Mild dementia samples from frontal pole with CDR equal to 2 or 3
2. FP.ND: No dementia samples from frontal pole with CDR less than 2
3. FP.SD: Severe dementia samples from frontal pole with CDR grater than 3
4. IFG.MD: Mild dementia samples from inferior frontal gyrus with CDR equal to 2 or 3
5. IFG.ND: No dementia samples from inferior frontal gyrus with CDR less than 2
6. IFG.SD: Severe dementia samples from inferior frontal gyrus with CDR grater than 3
7. PHG.MD: Mild dementia samples from para hippocampal gyrus with CDR equal to 2 or 3
8. PHG.ND: No dementia samples from para hippocampal gyrus with CDR less than 2
9. PHG.SD: Severe dementia samples from para hippocampal gyrus with CDR grater than 3
10. STG.MD: Mild dementia samples from superior temporal gyrus with CDR equal to 2 or 3
11. STG.ND: No dementia samples from superior temporal gyrus with CDR less than 2
12. STG.SD: Severe dementia samples from superior temporal gyrus with CDR grater than 3

Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp, fig.height=15, fig.width=13}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable, postAdjustCovars), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

@@

# Fit contrast
cntr = c()
for (region in c('FP', 'STG', 'PHG', 'IFG')){
  tmp = combn(paste0('BrainRegion.Diagnosis', region, '.', c('ND','MD','SD')),2)
  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
}

@@

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
                Study = 'MSSM',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('Region','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('Region','against'), sep = '\\.') %>%
  tidyr::unite(Comparison, reference, against, sep = '-') %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction, Region) %>%
  group_by(Region, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2) 
kable(tmp)

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red'))
p = p + facet_grid(Region+.~Comparison, scales = 'fixed')
p
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=15, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE %>% dplyr::filter(Comparison == 'ND-SD'), 
                 aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(Region~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE %>% dplyr::filter(Comparison == 'ND-SD'), 
                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(Region~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE %>% dplyr::filter(Comparison == 'ND-SD'), 
                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(Region~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

@@

### SVA Adjustments for eQTL analysis
Conditioned on primary variable (BrainRegion.Diagnosis) and identified covariates estimate surrogate variables using SVA package
```{r sva.adjust}
# Get (null) design matrix
MODEL0 = DESIGN$design[,-grep(primaryVariable,colnames(DESIGN$design))] # Get null model by removing variable of interest
MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
MODEL1 = DESIGN$design
MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]

# Source modified functions from sva package
# source('../R/num.sv.R')
source('../R/irwsva.build.R')
source('../R/f.pvalue.R')

# Calculate the number of surrogate variables
# NUM.SV = num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
#                      MODEL1, method = 'be', seed = 123456, B = 30, tol = 1e-20)

>>>>

knit2synapse::knitToFolderEntity(file = "./MSSM_REPROCESSED.Rmd",
                                 parentId = "syn8672415",
                                 entityName = 'MSSM Reprocessed RNASeq (Normalised)')
```

@@

```{r synapse.parameters, include=FALSE}
# Synapse parameters
parentId = 'syn8672415'
activityName = 'Covariate analysis';
activityDescription = 'Covariate analysis of MSSM reprocessed RNASeq data';

@@

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='AMPAD_reprocessing')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data}
# Get reprocessed sample counts for temporal cortex
COUNT_ID <- 'syn8691099';#'syn7994853';
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F, row.names = 1)

@@

# Get picard metrics from synapse
METADATA.PICARD_ID <- 'syn8698270';#'syn8380379'
ALL_USED_IDs[length(ALL_USED_IDs)+1] = METADATA.PICARD_ID
# METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = FALSE, exdir = getwd())
# METADATA.PICARD_FPATHS <- untar(synGet(METADATA.PICARD_ID)@filePath, list = TRUE, exdir = getwd())
# METADATA.PICARD_FPATHS = METADATA.PICARD_FPATHS[grep('picard.CombinedMetrics.csv', METADATA.PICARD_FPATHS)]
# METADATA.PICARD <- plyr::ldply(METADATA.PICARD_FPATHS, read.csv)
# METADATA.PICARD$sampleIdentifier = gsub('picard/', '', METADATA.PICARD_FPATHS) %>%
#   gsub('.accepted_hits.sort.coord/picard.CombinedMetrics.csv','',.)

METADATA.PICARD <- CovariateAnalysis::downloadFile(METADATA.PICARD_ID) %>%
  dplyr::rename(sampleIdentifier = sample)
colnames(METADATA.PICARD) = gsub('AlignmentSummaryMetrics__','',colnames(METADATA.PICARD))
colnames(METADATA.PICARD) = gsub('RnaSeqMetrics__','',colnames(METADATA.PICARD))

@@

# Fix metadata
METADATA = METADATA %>%
  dplyr::mutate(AOD = gsub('\\+','',AOD),
                BrainRegion = factor(BrodmannArea,
                                     levels = c('BM10', 'BM22', 'BM36', 'BM44'),
                                     labels = c('BM10' = 'FP', 'BM22' = 'STG', 
                                                'BM36' = 'PHG', 'BM44' = 'IFG')),
                Diagnosis = 'NONE', 
                RIN2 = RIN^2)

METADATA$Diagnosis = 'OTHER'
METADATA$Diagnosis[METADATA$CDR <= 0.5 & METADATA$bbscore <= 3 & METADATA$NP.1 <= 1] = 'CONTROL'
METADATA$Diagnosis[METADATA$CDR >= 1 & METADATA$bbscore >= 4 & METADATA$NP.1 >= 2] = 'AD'

levels(METADATA$batch)[levels(METADATA$batch) == ''] = 'NoBatch'

@@

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering}
primaryVariable <- c("CDR", "BrainRegion.Diagnosis")
FactorCovariates <- c("individualIdentifier", "batch", "RACE", "SEX", "BrainRegion.Diagnosis", "CDR")
ContCovariates <- c("RIN", "RIN2", "AOD","PMI",  "PCT_PF_READS_ALIGNED", "PCT_CODING_BASES",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES")

@@

PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(COUNT[genesToAnalyze, ], 
                                                 MIN_GENE_CPM=0, 
                                                 MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
```

### Outlier Analysis
```{r outlier.analysis, cache = FALSE, fig.width = 10}
indToRemove = c('hB_RNA_10622')

# Find principal components of expression to plot
PC <- prcomp(voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E, scale.=T, center = T)

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  dplyr::mutate(label = SampleID) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')
plotdata$label[!(plotdata$SampleID %in% indToRemove)] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Diagnosis, size=RIN, shape = SEX))
p <- p + theme_bw() + theme(legend.position="right")
p <- p + geom_text(aes(label= label), size=4, hjust=0)
p

# Plot abberent distribution of logcpm counts
tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
  rownameToFirstColumn('Gene.ID') %>%
  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')

p = ggplot(tmp1 %>%
             dplyr::filter(SampleID %in% indToRemove),
           aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'top') + facet_grid(Diagnosis+.~BrainRegion)
p

indToRetain = setdiff(colnames(PROCESSED_COUNTS$filteredExprMatrix$counts), indToRemove)
PROCESSED_COUNTS$filteredExprMatrix$counts = PROCESSED_COUNTS$filteredExprMatrix$counts[,indToRetain]
COVARIATES = COVARIATES[indToRetain,]

tmp = COVARIATES %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
  group_by(BrainRegion, Diagnosis) %>%
  summarise(count = n()) %>%
  spread(Diagnosis, count)
```
Processing `r dim(PROCESSED_COUNTS$filteredExprMatrix)[1]` genes in `r dim(PROCESSED_COUNTS$filteredExprMatrix)[2]` samples

Based on the expression pattern following samples were tagged as outliers: `r paste(indToRemove, collapse = ', ')`

Distribution of samples are: `r kable(tmp)`

### Conditional Quantile Normalisation (CQN)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

@@

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```

```{r explore.data, cache=FALSE, fig.height=5, fig.width=10}
# Plot abberent distribution of logcpm counts
tmp1 = (CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset) %>%
  rownameToFirstColumn('Gene.ID') %>%
  tidyr::gather(SampleID, logCPM, -Gene.ID) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID')) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.')

p = ggplot(tmp1, aes(x = logCPM, color = SampleID)) + geom_density() 
p = p + theme(legend.position = 'NONE') + facet_grid(BrainRegion+.~Diagnosis, scale = 'free')
p
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Decomposition of expression
Clustering of library normalised unadjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=5, fig.width=10, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Diagnosis, size=RIN, shape = SEX))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(.~Region, scales = 'free_y')
p
```
Tree based classification of samples
```{r decompse.normalise.data1.2, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = (COVARIATES[,c(FactorCovariates), drop = F])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residual.adj, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

@@

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Diagnosis, size=RIN, shape = SEX))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(.~Region, scales = 'free_y')
p
```

@@

### Differential expression analysis (with BrainRegion.CDR as primary variable)
Differential expression is performed on the primary variable by controlling for covariates identified above

Interpretation of BrainRegion.CDR
1. FP.CDR: change in CDR in frontal pole where 0 no cognitive decline and 5 severe cognitive decline
2. IFG.CDR: change in CDR in inferior frontal gyrus where 0 no cognitive decline and 5 severe cognitive decline
3. PHG.CDR: change in CDR in parahippocampal gyrus where 0 no cognitive decline and 5 severe cognitive decline
4. STG.CDR: change in CDR in superior temporal gyrus where 0 no cognitive decline and 5 severe cognitive decline

Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp, fig.height=6, fig.width=15}
# Get design matrix
COVARIATES.tmp = COVARIATES %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Diagnosis'), sep = '\\.') %>%
  dplyr::mutate(BrainRegion = factor(BrainRegion))
DESIGN = getDesignMatrix(COVARIATES.tmp[, c('BrainRegion', postAdjustCovars), drop = F], Intercept = F)
ind = grep('BrainRegion', colnames(DESIGN$design))
DESIGN$design[,ind] = DESIGN$design[,ind] * as.numeric(COVARIATES.tmp[,'CDR'])
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

@@

# Fit contrast
contrast = makeContrasts(contrasts = c("BrainRegionFP", "BrainRegionIFG", "BrainRegionPHG", "BrainRegionSTG"),
                         levels = colnames(FIT$coefficients))
FIT.CONTR = contrasts.fit(FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:4, function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = T) %>%
    rownameToFirstColumn('ensembl_gene_id')
}, FIT.CONTR)
names(DE) = colnames(contrast)

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion','',Comparison),
                Study = 'MSSM',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  dplyr::rename(BrainRegion = Comparison) %>%
  dplyr::mutate(Comparison = 'CDR') %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion) %>%
  group_by(BrainRegion, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2) 
kable(tmp)

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(Comparison+.~BrainRegion, scales = 'free')
p
```

METADATA$Diagnosis[METADATA$CDR <= 0.5 & METADATA$bbscore <= 3 & METADATA$NP.1 <= 1] = 'CONTROL'
METADATA$Diagnosis[METADATA$CDR >= 1 & METADATA$bbscore >= 4 & METADATA$NP.1 >= 2] = 'AD'

Interpretation of BrainRegion.Diagnosis
1. FP.AD: Samples from frontal pole region with CDR >= 1 and bbscore >= 4 and NP.1 >= 2
2. FP.CONTROL: Samples from frontal pole region with CDR <= 0.5 and bbscore <= 3 and NP.1 <= 1
3. FP.OTHER: All the other frontal pole samples
4. PHG.AD: Samples from parahippocampal gyrus region with CDR >= 1 and bbscore >= 4 and NP.1 >= 2
5. PHG.CONTROL: Samples from parahippocampal gyrus region with CDR <= 0.5 and bbscore <= 3 and NP.1 <= 1
6. PHG.OTHER:All the other parahippocampal gyrus samples
7. STG.AD: Samples from frontal superior temporal gyrus with CDR >= 1 and bbscore >= 4 and NP.1 >= 2
8. STG.CONTROL: Samples from superior temporal gyrus region with CDR <= 0.5 and bbscore <= 3 and NP.1 <= 1
9. STG.OTHER: All the other superior temporal gyrus samples
10. IFG.AD: Samples from inferior frontal gyrus region with CDR >= 1 and bbscore >= 4 and NP.1 >= 2
11. IFG.CONTROL: Samples from inferior frontal gyrus region with CDR <= 0.5 and bbscore <= 3 and NP.1 <= 1
12. IFG.OTHER: All the other inferior frontal gyrus samples

Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp1, fig.height=15, fig.width=13}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable[2], postAdjustCovars), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

# Fit contrast
cntr = c()
for (region in c('FP', 'STG', 'PHG', 'IFG')){
  tmp = combn(paste0('BrainRegion.Diagnosis', region, '.', c('OTHER','AD','CONTROL')),2)
  cntr = c(cntr, apply(tmp, 2, paste, collapse = '-'))
}

@@

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('BrainRegion.Diagnosis','',Comparison),
                Study = 'MSSM',
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, c('from.state','to.state'), sep = '-') %>%
  tidyr::separate(from.state, c('BrainRegion','reference'), sep = '\\.') %>%
  tidyr::separate(to.state, c('BrainRegion','against'), sep = '\\.') %>%
  tidyr::unite(Comparison, reference, against, sep = '-') %>%
  left_join(GENE.PARAM %>%
              dplyr::select(ensembl_gene_id, hgnc_symbol, percentage_gc_content, gene.length) %>%
              unique())
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::select(ensembl_gene_id, Comparison, Direction, BrainRegion) %>%
  group_by(BrainRegion, Comparison, Direction) %>%
  dplyr::summarise(FDR_0_05_FC_1.2 = length(unique(ensembl_gene_id))) %>%
  spread(Direction, FDR_0_05_FC_1.2) 
kable(tmp)

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position='top')
p = p + facet_grid(BrainRegion+.~Comparison, scales = 'fixed')
p
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=15, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE %>% dplyr::filter(Comparison == 'AD-CONTROL'), 
                 aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(BrainRegion~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE %>% dplyr::filter(Comparison == 'AD-CONTROL'), 
                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(BrainRegion~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE %>% dplyr::filter(Comparison == 'AD-CONTROL'), 
                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(BrainRegion~.+Comparison) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

@@

### SVA Adjustments for eQTL analysis
Conditioned on primary variable (BrainRegion.Diagnosis) and identified covariates estimate surrogate variables using SVA package
```{r sva.adjust}
# Get (null) design matrix
MODEL0 = DESIGN$design[,-grep(primaryVariable[2],colnames(DESIGN$design))] # Get null model by removing variable of interest
MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
MODEL1 = DESIGN$design
MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]

# Source modified functions from sva package
source('../R/irwsva.build.R')
source('../R/f.pvalue.R')
